\textbf{Political Instability and Financial Markets}

### 2017-08-23 doc/paper.tex
Made changes to conclusion and section 4. Also added sub-subsections to section 3. I actually think we should make section 4 a subsection of section 3, but we can discuss before I do that.

\singlespacing
\begin{abstract}
We examine the relationship between political instability and daily returns of national stock indices. A common hypothesis is that political instability discourages investment and reduces economic growth. By contrast, we find that while financial volatility increases dramatically following ``irregular'' regime changes, different types of regime changes have disparate effects on stock returns. We use daily financial data and a constant mean return event study model to show that abnormal returns following resignations are large and positive (6\%), while abnormal returns following assassinations are negative and smaller in magnitude (1\%). The impact of coups tends to be negative, but some events such as the 2002 temporary ousting of Hugo Chavez in Venezuela result in positive abnormal returns of 20\% or more. We therefore find that the type of political event and its expected impact on economic policy determines the direction of abnormal returns.
\end{abstract}
\doublespacing

@@

\section{Irregular Regime Changes} \label{sec: Irregular Regime Changes}
\subsection{Abnormal Returns} \label{subsec: Abnormal Returns}
We begin our analysis by studying the effect of irregular regime changes on stock returns. We follow the standard event study methodology as presented by, among others, \citep{mackinlay1997event} and \citep{campbell1997econometrics}. Normal performance is measured with a constant mean return model\footnote{A constant mean return model is used instead of a market model in order to maximize the number of observations (plausible market indices such as the MSCI World Index and the S\&P/IFC Emerging Markets Investable Composite Index only begin in 1976 and 1995 respectively). That said, results are insensitive to the use of a market model.},
\begin{align} \label{eqn:market-model}
R_{it}=\mu_{i}+\epsilon_{it},
\end{align}
where $R_{it}$ is the logged return of national stock index $i$ on trading day $t$ and $\epsilon_{it}$ is the error term. We calculate abnormal returns (ARs), in an ``event window'' surrounding the date of each coup, $AR_{i\tau}=R_{i\tau}-\widehat{\mu}_i$, where $\tau$ is a date in the event window, and $\widehat{\mu}_i$ is estimated in an ``estimation window'' preceding the event window with \autoref{eqn:market-model}. We use a 41 day event window (i.e. 20 pre-event trading days, the event day, and 20 post-event trading days). The estimation window is the 250 trading days prior to the start of the event window. The abnormal returns are then used to generate cumulative abnormal returns (CARs) between event day $\tau_1$ and event day $\tau_2$: $CAR(\tau_1,\tau_2)=\sum_{\tau=\tau_1}^{\tau_2}AR_{i\tau}$.

@@

We report abnormal returns separately for coups, assassinations and resignations because they may have distinct effects on stock returns. \autoref{tab:AR-coups} shows abnormal returns for national stock indices both preceding and following coup d'etat. Standard errors and p-values are calculated using asymptotic t-statistics as in \citep{mackinlay1997event}.\footnote{It is appropriate to use the standard normal distribution to calculate test statistics because the length of the estimation window is sufficiently long (250 trading days).} \autoref{tab:AR-coups} contains all coups presented in \autoref{tab:coup list} with the exception of the partial/failed Venezuelan coup attempt of 2002. We examine this event as well, but do not include it in the sample in \autoref{tab:AR-coups} as it does not match our definition of a successful coup.

@@

Event day ARs are negative for all coups except the 1971 coup in Argentina and the 2002 failed coup d'etat attempt against Hugo Chavez (\autoref{fig:AR-Ven}). These results provide evidence that coups can in fact result in positive abnormal returns. While the 1971 Argentinian coup did result in another military leader, it did so while calling for free and democratic elections and replaced a government that had adopted extreme protectionist economic policies. The ultimately failed coup against Hugo Chavez replaced a left-wing populist government with a new pro-business president.

@@

@@

These results are consistent with our hypothesis that the nature of the political event and its expect impact on policy matters. While the mean effect of assassinations is negative, it is smaller in magnitude than for coups. Unlike a coup, an assassination is typically a random event that will not necessarily cause immediate change in economic policy. As such, we would expect CARs to be negative due to increased instability and uncertainty, but smaller in magnitude to a coup or resignation due to greater expectations of policy inertia.

@@

@@

While the Egyptian Revolution may have positive pro-democracy effects on the Egpytian political system, its short-term impact on the economy was disastrous. As shown in \autoref{fig:CAR-Egypt}, abnormal returns on the Egyptian Stock Exchange Index (EGX 30) were around -7\% on January 26th and -10\% the day after. To prevent further decline during the uprising, the Egyptian Stock Exchanged closed at the end of trading on January 27th. President Mubarak resigned on February 11, but the market remained closed until March 23, when CARs declined by another 9\%, before rebounding slightly thereafter.

@@

The start and end dates in \autoref{tab:protest-list} are the dates that protests began and leader's resigned respectively. Resignations caused by popular uprisings were identified by examining the descriptions in the Coup d'?tat Events Handbook and Archigos Version 2.9. Additional Lexis Nexis searches were used to verify these descriptions. Start dates were determined through similar means.

In \autoref{tab:protest-stocks}, we examine whether public protests influence stock prices. The variable, \textit{Protest} is equal to 1 during the dates in which citizens participate in political activities demanding the resignation of the executive are and 0 otherwise. Non-protest dates are the 250 days prior to the start dates and after the end dates listed in \autoref{tab:protest-list}.\footnote{The volatility estimates used as the dependent variable in column (4) are estimated on the 250 days prior to the start date, the protest dates, and the 250 days following the end date.}

@@

\begin{table}[!ht]
\caption{Effect of public protests on stock prices} \label{tab:protest-stocks}
\vspace{-5pt}
\footnotesize
\begin{center}
\begin{threeparttable}
\begin{tabular*}{\textwidth}{l@{\extracolsep{\fill}}d{1.3}d{1.3}d{1.3}d{1.3}}
  \hline
  \hline
\multicolumn{1}{c}{}&\multicolumn{1}{c}{Returns} & \multicolumn{2}{c}{Absolute Value of Returns}&\multicolumn{1}{c}{Volatility}\\
\cmidrule(r){2-2} \cmidrule(r){3-4} \cmidrule(r){5-5}
 & \multicolumn{1}{c}{(1)}&\multicolumn{1}{c}{(2)}&\multicolumn{1}{c}{(3)}&\multicolumn{1}{c}{(4)}\\
  \hline
\ExpandableInput{../tables/protest-regtable.txt}
   \hline
   \hline
\end{tabular*}
\scriptsize
Notes: Standard errors clustered by event are in parentheses. Statistical significance at the 10\%, 5\%, and 1\% level is denoted by *, ** and *** respectively.
\end{threeparttable}
\end{center}
\end{table}

\section{Conclusion}
This paper has examined the economic effects of irregular regime changes. It uses an event study approach to show that investors expect irregular regime changes to have large effects on equity returns. This methodology is less susceptible to endogeneity biases than studies that use cross country data.

The results are consistent with the idea that government competence is a random walk and that government changes have large impacts on investor confidence. Financial volatility surrounding regime changes, which is often characterized by large positive \textit{and} negative stock returns, suggests that irregular regime changes increase policy uncertainty. The variation in the direction of abnormal returns is consistent with the idea that not all governments are created equally. Abnormal returns are likely positive following resignations because those leaders were the most likely to be ``bad'' leaders.

>>>>

\singlespacing
\begin{abstract}
We examine the relationship between political instability and daily returns of national stock indices. A common hypothesis is that political instability discourages investment and reduces economic growth. By contrast, we find that while financial volatility increases dramatically following ``irregular'' regime changes, different types of regime change have disparate effects on stock returns. We use daily financial data and a constant mean return event study model to show that abnormal returns following resignations are large and positive (6\%), while abnormal returns following assassinations are negative and smaller in magnitude (1\%). The impact of coups tends to be negative, but some events result in positive abnormal returns of 20\% or more. We also find that volatility increases during times of protest preceding resignations, but that no clear directionality is present. We therefore find that the type of political event and its expected impact on economic policy determines the direction of abnormal returns.
\end{abstract}
\doublespacing

@@

\section{Impact of Political Instability on Stock Returns} \label{sec: Impact of Political Instability on Stock Returns}
\subsection{Abnormal Returns} \label{subsec: Abnormal Returns}
We begin our analysis by studying the effect of irregular regime changes on stock returns. We follow the standard event study methodology as presented by, among others, \citep{mackinlay1997event} and \citep{campbell1997econometrics}. Normal performance is measured with a constant mean return model\footnote{A constant mean return model is used instead of a market model in order to maximize the number of observations (plausible market indices such as the MSCI World Index and the S\&P/IFC Emerging Markets Investable Composite Index only begin in 1976 and 1995 respectively). That said, results are insensitive to the use of a market model.},
\begin{align} \label{eqn:market-model}
R_{it}=\mu_{i}+\epsilon_{it},
\end{align}
where $R_{it}$ is the logged return of national stock index $i$ on trading day $t$ and $\epsilon_{it}$ is the error term. We calculate abnormal returns (ARs), in an ``event window'' surrounding the date of each coup, $AR_{i\tau}=R_{i\tau}-\widehat{\mu}_i$, where $\tau$ is a date in the event window, and $\widehat{\mu}_i$ is estimated in an ``estimation window'' preceding the event window with \autoref{eqn:market-model}. We use a 41 day event window (i.e. 20 pre-event trading days, the event day, and 20 post-event trading days). The estimation window is the 250 trading days prior to the start of the event window. The abnormal returns are then used to generate cumulative abnormal returns (CARs) between event day $\tau_1$ and event day $\tau_2$: $CAR(\tau_1,\tau_2)=\sum_{\tau=\tau_1}^{\tau_2}AR_{i\tau}$.

@@

We report abnormal returns separately for coups, assassinations and resignations because they may have distinct effects on stock returns. \autoref{tab:AR-coups} shows abnormal returns for national stock indices both preceding and following coup d'etat. Standard errors and p-values are calculated using asymptotic t-statistics as in \citep{mackinlay1997event}.\footnote{It is appropriate to use the standard normal distribution to calculate test statistics because the length of the estimation window is sufficiently long (250 trading days).}

\subsubsection{Coups} \label{subsec: Coups}
\autoref{tab:AR-coups} contains all coups presented in \autoref{tab:coup list} with the exception of the partial/failed Venezuelan coup attempt of 2002. We examine this event as well, but do not include it in the sample in \autoref{tab:AR-coups} as it does not match our definition of a successful coup.

@@

Event day ARs are negative for all coups except the 1971 coup in Argentina and the 2002 failed coup d'etat attempt against Hugo Chavez (\autoref{fig:AR-Ven}). These results provide evidence that coups can in fact result in positive abnormal returns. While the 1971 Argentinian coup did result in another military leader, it did so while calling for free and democratic elections and replaced a government that had adopted extreme protectionist economic policies. By 1973 Argentina had transitioned to a democracy.\footnote{Based on Polity IV polity score of 6. Values of 6-10 are defined as democracies.} The ultimately failed coup against Hugo Chavez replaced a left-wing populist government with a new pro-business president.

@@

\subsubsection{Assassinations} \label{subsec: Assassinations}

@@

These results are consistent with our hypothesis that the nature of the political event and its expected impact on policy matters. While the mean effect of assassinations is negative, it is smaller in magnitude than for coups. Unlike a coup, an assassination is typically a random event that will not necessarily cause immediate change in economic policy. As such, we would expect CARs to be negative due to increased instability and uncertainty, but smaller in magnitude to a coup or resignation due to greater expectations of policy inertia.

@@

\subsubsection{Resignations} \label{subsec: Resignations}

@@

While the Egyptian Revolution may have had positive pro-democracy effects on the Egpytian political system, its short-term impact on the economy was disastrous. As shown in \autoref{fig:CAR-Egypt}, abnormal returns on the Egyptian Stock Exchange Index (EGX 30) were around -7\% on January 26th and -10\% the day after. To prevent further decline during the uprising, the Egyptian Stock Exchanged closed at the end of trading on January 27th. President Mubarak resigned on February 11, but the market remained closed until March 23, when CARs declined by another 9\%, before rebounding slightly thereafter.

@@

The start and end dates in \autoref{tab:protest-list} are the dates that protests began and leader's resigned respectively. Resignations caused by popular uprisings were identified by examining the descriptions in the Coup d'etat Events Handbook and Archigos Version 4.1. Additional Lexis Nexis searches were used to verify these descriptions. Start dates were determined through similar means.

In \autoref{tab:protest-stocks}, we examine whether public protests influence stock prices. The variable \textit{Protest} is equal to 1 during the dates in which citizens participate in political activities demanding the resignation of the executive and 0 otherwise. Non-protest dates are the 250 days prior to the start dates and after the end dates listed in \autoref{tab:protest-list}.\footnote{The volatility estimates used as the dependent variable in column (4) are estimated on the 250 days prior to the start date, the protest dates, and the 250 days following the end date.}

@@

We therefore find that both volatility and the absolute value of returns increase during times of protest. Similarly to coups, however, the direction of returns is dependent upon the nature of the protest in question.

\begin{table}[!ht]
\caption{Effect of public protests on stock prices} \label{tab:protest-stocks}
\vspace{-5pt}
\footnotesize
\begin{center}
\begin{threeparttable}
\begin{tabular*}{\textwidth}{l@{\extracolsep{\fill}}d{1.3}d{1.3}d{1.3}d{1.3}}
  \hline
  \hline
\multicolumn{1}{c}{}&\multicolumn{1}{c}{Returns} & \multicolumn{2}{c}{Absolute Value of Returns}&\multicolumn{1}{c}{Volatility}\\
\cmidrule(r){2-2} \cmidrule(r){3-4} \cmidrule(r){5-5}
 & \multicolumn{1}{c}{(1)}&\multicolumn{1}{c}{(2)}&\multicolumn{1}{c}{(3)}&\multicolumn{1}{c}{(4)}\\
  \hline
\ExpandableInput{../tables/protest-regtable.txt}
   \hline
   \hline
\end{tabular*}
\scriptsize
Notes: Standard errors clustered by event are in parentheses.
\end{threeparttable}
\end{center}
\end{table}

\section{Conclusion}
This paper has examined the effects of irregular regime changes on financial returns. It uses an event study approach to show that investors expect irregular regime changes to have large effects on equity returns. This methodology is less susceptible to endogeneity biases than studies that use cross country data.

The results are consistent with the idea that government competence is a random walk and that government changes have large impacts on investor confidence. Financial volatility surrounding regime changes, which is often characterized by large positive \textit{and} negative stock returns, suggests that irregular regime changes increase policy uncertainty. The variation in the direction of abnormal returns is consistent with the idea that not all types of regime changes are equivalent in terms of expected economic policy outcomes. Abnormal returns are likely positive following resignations because those leaders were likely to be ``bad'' leaders, and coups may result in positive returns in the event of a transition to more democratic or pro-business leadership.
